`conmat`: generate synthetic contact matrices for a given age-stratified population

### 2024-08-19 paper/paper.qmd

---
title: 'conmat: generate synthetic contact matrices for a given age-stratified population'
authors:
- affiliation: 1
  name: Nicholas Tierney
  orcid: 0000-0003-1460-8722
- affiliation: 1,2
  name: Nick Golding
  orcid: 0000-0001-8916-5570
- affiliation: 1,3
  name: Aarathy Babu
  orcid: 
- affiliation: 4
  name: Michael Lydeamore
  orcid: 0000-0001-6515-827X
- affiliation: 1,3
  name: Chitra Saraswati
  orcid: 0000-0002-8159-0414
date: today
bibliography: references.bib
tags:
- epidemiology
- R
- infectious disease
affiliations:
- index: 1
  name: Telethon Kids Institute
- index: 2
  name: Curtin University
- index: 3
  name: 
- index: 4
  name: Monash University
execute: 
  echo: true
  cache: true
format: 
  pdf: 
    keep-md: true
    fig-height: 4
    fig-align: center
    fig-format: png
    dpi: 300
  html: 
    keep-md: true
    fig-height: 4
    fig-align: center
    fig-format: png
    dpi: 300
---

@@

These social contact surveys can be projected on to a given demographic structure to produce estimated daily contact rates between age groups. These are known as "contact matrices" or "synthetic contact matrices". A widely used approach by @prem2021 [@prem2021] produced contact matrices for 177 countries at "urban" and "rural" levels for each country.

@@

`conmat` was built to predict at four settings: work, school, home, and other. 
One model is fitted for each setting. 
Each model fitted is a Poisson generalised additive model (GAM) which predicts the count of contacts, with an offset for the log of participants. 
The model has six covariates to explain six key features of the relationship between ages, 
and two optional covariates for attendance at school or work.
The two optional covariates are included depending on which setting the model is fitted for.

[njt: should we mention the link/family?]

Each cell in the resulting contact matrix, indexed ($i$, $j$), is the predicted number of people in age group $j$ that a single individual in age group $i$ will have contact with per day. The sum over all of the $j$ age groups for a particular age group $i$ is the predicted total number of contacts per day for each individual of age group $i$.

[njt: do we need to mention that we need to do the exponentiated sum, since these are on the log scale?]

@@

[njt: OK with putting these as dot points?]

@@

plot_age_term_settings <- gg_age_terms_settings(age_predictions_all_settings)
age_predictions_long <- pivot_longer_age_preds(age_predictions)
plot_age_predictions_long <- gg_age_partial_pred_long(age_predictions_long) +
  coord_equal() +
  labs(x = "Age from",
       y = "Age to") + 
  theme(
    legend.position = "bottom"
  ) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  expand_limits(x = c(0, 100), y = c(0, 100))

@@

```{r}
#| label: show-partial-plots
#| echo: false
#| fig-height: 8
#| fig-cap: "Partial predictive plot (A) and overall synthetic contact matrix (B) for the Poisson GAM fitted to the POLYMOD contact survey in the home setting. The strong diagonal elements, and parents/grandparents interacting with children result in the classic 'diagonal with wings' shape."
plot_all_terms_sum <- plot_age_predictions_long +
  plot_age_predictions_sum + 
  plot_layout(design = "
            AAAABBBB
            AAAABBBB
            AAAABBBB
            AAAABBBB
            ") + 
  plot_annotation(
    tag_levels = "A"
  )

@@

The covariates used by conmat were designed to represent the key features that are typically present in a contact matrix for different settings (work, school, home, other). Including other sources of information that may better describe these contact patterns, such as inter-generational mixing, or differences in school ages of a local demographic, may improve model performance.

The interface to the model formula in conmat is fixed; users cannot change the covariates of the model. This means that if there is an unusual structure in their contact data it might not be accurately captured by conmat. It was a design decision that was made to focus on the key feature of conmat: using just age population data to predict a contact matrix.

>>>>

---
title: '`conmat`: generate synthetic contact matrices for a given age-stratified population'
authors:
- affiliation: 1
  name: Nicholas Tierney
  orcid: 0000-0003-1460-8722
- affiliation: 1,2
  name: Nick Golding
  orcid: 0000-0001-8916-5570
- affiliation: 1,3
  name: Aarathy Babu
  orcid: 
- affiliation: 4
  name: Michael Lydeamore
  orcid: 0000-0001-6515-827X
- affiliation: 1,3
  name: Chitra Saraswati
  orcid: 0000-0002-8159-0414
date: today
bibliography: references.bib
cite-method: biblatex
tags:
- epidemiology
- R
- infectious disease
affiliations:
- index: 1
  name: Telethon Kids Institute
- index: 2
  name: Curtin University
- index: 3
  name: 
- index: 4
  name: Monash University
execute: 
  echo: true
  cache: false
format: 
  pdf: 
    keep-md: true
    fig-height: 4
    fig-align: center
    fig-format: png
    dpi: 300
  html: 
    keep-md: true
    fig-height: 4
    fig-align: center
    fig-format: png
    dpi: 300
---

@@

These social contact surveys can be projected on to a given demographic structure to produce estimated daily contact rates between age groups. These are known as "contact matrices" or "synthetic contact matrices". A widely used approach by @prem2017 [@prem2021] produced contact matrices for 177 countries at "urban" and "rural" levels for each country.

@@

`conmat` was built to predict at four settings: work, school, home, and other. 
One model is fitted for each setting. 
Each model fitted is a Poisson generalised additive model (GAM) with a log link function, which predicts the count of contacts, with an offset for the log of participants. 
The model has six covariates to explain six key features of the relationship between ages, 
and two optional covariates for attendance at school or work.
The two optional covariates are included depending on which setting the model is fitted for.

Each cell in the resulting contact matrix (after back-transformation of the link function), indexed ($i$, $j$), is the predicted number of people in age group $j$ that a single individual in age group $i$ will have contact with per day. The sum over all of the $j$ age groups for a particular age group $i$ is the predicted total number of contacts per day for each individual of age group $i$.

@@

@@

plot_age_term_settings <- gg_age_terms_settings(age_predictions_all_settings)
age_predictions_long <- pivot_longer_age_preds(age_predictions)
plot_age_predictions_long <- gg_age_partial_pred_long(age_predictions_long) +
  coord_equal() +
  labs(
    x = "Age from",
    y = "Age to"
    ) + 
  theme(
    legend.position = "bottom",
    axis.text = element_text(size = 6),
    panel.spacing = unit(x = 1, units = "lines")
  ) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  expand_limits(x = c(0, 100), y = c(0, 100))

@@

```{r}
#| label: show-partial-plots
#| echo: false
#| fig-cap: "Partial predictive plot (A) and overall synthetic contact matrix (B) for the Poisson GAM fitted to the POLYMOD contact survey in the home setting. The strong diagonal elements, and parents/grandparents interacting with children result in the classic 'diagonal with wings' shape."
plot_all_terms_sum <- plot_age_predictions_long +
  plot_age_predictions_sum + 
  plot_layout(design = "
            AAAAABBBB
            AAAAABBBB
            AAAAABBBB
            AAAAABBBB
            ") + 
  plot_annotation(
    tag_levels = "A"
  )

@@

The covariates used by `conmat` were designed to represent the key features that are typically present in a contact matrix for different settings (work, school, home, other). Including other sources of information that may better describe these contact patterns, such as inter-generational mixing, or differences in school ages of a local demographic, may improve model performance.

The interface to the model formula in `conmat` is fixed; users cannot change the covariates of the model. This means if there is an unusual structure in their contact data it might not be accurately captured by `conmat`. This fixed formula was a design decision made to focus on the key feature of `conmat`: using just age population data to predict a contact matrix.
